Relative Stable Maps in Genus One via Central Alignments

### 2018-11-30 new.tex
started the fight against the boundary

@@

\begin{cor}
 $\VZ_{1,\alpha}(\PP^N|H,d)$ is smooth over its Artin fan.
\end{cor}

@@

\section{An explicit description of Gathmann's boundary}

\subsection{The line bundle and section}
On $\VZrel{1}{\alpha}{X}{\beta}$, as on any space of maps to an expansion really, there is not only a universal curve $C$ with universal map $f$ to the expansion $W$, but also the stabilisation of the collapsed map obtained by composing $f$ with $\pi_X\colon W\to X$:
\bcd
C\ar[r,"f"]\ar[d,"\stab"] & W \ar[d,"\pi_X"] \\
\stC\ar[r,"\stf"] & X
\ecd
Notice that the image of the marking $p_k$ on $\stC$ has either contact order $\alpha_k$ to $Y$, or it lies on an internal component. The section $s\in\Gamma(X,\OO_X(Y))$ whose vanishing locus defines $Y$, induces a section $\sts$ of $\stpi_*\stf^*\OO_X(Y)$. By the previous discussion, the image of $\sts$ in $\ev_k^*\mathcal P^{\alpha_k}\OO_X(Y)$ vanishes, hence by the principal parts exact sequence
\[ 0\to p_k^*\Omega_{\stC}^{\otimes\alpha_k+1}\otimes\ev_k^*\OO_X(Y)\to \ev_k^*\mathcal P^{\alpha_k+1}\OO_X(Y)\to \ev_k^*\mathcal P^{\alpha_k}\OO_X(Y)\to 0\]
we obtain a section $\sts_k$ of the line bundle $p_k^*\Omega_{\stC}^{\otimes\alpha_k+1}\otimes\ev_k^*\OO_X(Y)$.

It is apparent that, set-theoretically at least, the vanishing locus of $\sts_k$ consists of those maps such that $p_k$ lies on an internal component of $\stC$, or, equivalently, it is mapped to higher level in the expansion and lies not on a ``trivial bubble'' in $C$.

>>>>

\def\Yagraph{\tikz[baseline=-3pt,scale=.8]{
\draw (2,0) -- (0,1) (2,0) -- (0,.5) (2,0) -- (0,-1);
\draw (2,0) circle(2pt)[fill=black];
\draw (0,1) circle(2pt)[fill=white];
\draw (0,.5) circle(2pt)[fill=black];
\draw (0,-.5) node{$\ldots$};
\draw (0,-1) circle(2pt)[fill=black];
\draw (-1,1.5) -- (1,1.5) -- (1,-1.5) -- (-1,-1.5) -- (-1,1.5);
\draw (1,1.5) -- (3,2) -- (3,-1) -- (1,-1.5);
}}

\def\Ybgraph{\tikz[baseline=-3pt,scale=.8]{
\draw (2,0) to[out=120,in=0] (0,1) (2,0) -- (0,1) (2,0) -- (0,.5) (2,0) -- (0,-1);
\draw (2,0) circle(2pt)[fill=black];
\draw (0,1) circle(2pt)[fill=black];
\draw (0,.5) circle(2pt)[fill=black];
\draw (0,-.5) node{$\ldots$};
\draw (0,-1) circle(2pt)[fill=black];
\draw (-1,1.5) -- (1,1.5) -- (1,-1.5) -- (-1,-1.5) -- (-1,1.5);
\draw (1,1.5) -- (3,2) -- (3,-1) -- (1,-1.5);
}}

\def\Ycgraph{\tikz[baseline=-3pt,scale=.8]{
\draw (2,0) -- (0,1) (2,0) -- (0,.5) (2,0) -- (0,-1);
\draw (2,0) circle(2pt)[fill=white];
\draw (0,1) circle(2pt)[fill=black];
\draw (0,.5) circle(2pt)[fill=black];
\draw (0,-.5) node{$\ldots$};
\draw (0,-1) circle(2pt)[fill=black];
\draw (-1,1.5) -- (1,1.5) -- (1,-1.5) -- (-1,-1.5) -- (-1,1.5);
\draw (1,1.5) -- (3,2) -- (3,-1) -- (1,-1.5);
}}

@@

\begin{cor}\label{cor:log_smooth}
 $\VZ_{1,\alpha}(\PP^N|H,d)$ is smooth over its Artin fan.
\end{cor}

@@

\section{Description of the boundary}

Note that $f$ has contact order with $H$ exactly equal to $\alpha_k$ at the marking $p_k$, or else the irreducible component of $C$ on which $x_k$ lies is mapped entirely inside $H$. By pulling back along $f$ the equation defining $H$, and taking its $\alpha_k+1$-st derivative at $x_k$ (which makes sense because all the lower order derivatives do vanish by assumption) we single out the latter locus. By staring at the exact sequence of jet bundles
\[0\to p_k^*\Omega_{C}^{\otimes\alpha_k}\otimes\ev_k^*\OO_{\PP^N}(H)\to p_k^*\mathcal J^{\alpha_k}(f^*\OO_{\PP^N}(H))\to p_k^*\mathcal J^{\alpha_k-1}(f^*\OO_{\PP^N}(H))\to 0\]
we realise that there is on $\VZ_{1,\alpha}(\PP^N|H,d)$ a line bundle (the leftmost term in the exact sequence) with a natural section cutting the locus where the curve breaks, and the piece containing $p_k$ is mapped inside $H$.
\begin{lem}
 \[(\alpha_k\psi_k+\ev_k^*H)[\VZ_{1,\alpha}(\PP^N|H,d)]=[D_{1,\alpha;k}(\PP^N|H,d)],\]
\end{lem}
In what follows we shall give an explicit (recursive) description of the right-hand side, in terms of tautological classes on spaces of maps with lower numerical invariants.
\begin{remark}
 The divisors we are after are a union of logarithmic strata, because the locus where the log structure is trivial coincides precisely with the nice locus.
\end{remark}
By Corollary \ref{cor:log_smooth}, the divisorial logarithmic strata correspond to rays of the Artin fan.

\subsection{Combinatorial description}
As long as the minimal subcurve of genus one has positive degree, the picture is classical: the dual graph of the curve is bipartite - vertices mapped inside $H$ are distinguished from the others. In fact there is only one such vertex, because, if there were more, the relative position of their image in $\mathbb R_{\geq 0}$ under the tropical map would be unconstrained, therefore spanning a locus larger than a ray in the tropical moduli space. We follow in R. Vakil's footsteps in setting the following notation:
\begin{figure}[h]
    \centering
    \begin{minipage}{0.3\textwidth}
        \centering
        \Yagraph
        \caption{$\mathcal Y_a$}
    \end{minipage}\hfill
    \begin{minipage}{0.3\textwidth}
        \centering
        \Ybgraph
        \caption{$\mathcal Y_b$}
    \end{minipage}\hfill
    \begin{minipage}{0.3\textwidth}
        \centering
        \Ycgraph
        \caption{$\mathcal Y_c^+$}
    \end{minipage}
\end{figure}

Notice that continuity of the tropical map determines the mutual relationship among the edge lengths.

On the other hand, when there is a contracted elliptic subcurve - and it will be contracted into the hyperplane, because otherwise it wouldn't be generic, by density of the nice locus in $\VZ_{1,\tilde\alpha}(\PP^N|H,\tilde d)$ - the picture becomes much more complicated due to the alignment. The combs may break multiple times and they may even point in opposite directions.
\begin{figure}
 \tikz{
 \draw (0,0) circle(2pt)[fill=black] (0,2) circle(2pt)[fill=black] (0,4) circle(2pt)[fill=black] (0,6) circle(2pt)[fill=black] (0,-2) circle(2pt)[fill=black] (2,0) circle(2pt)[fill=white] (1.1,1.78) circle(2pt)[fill=black] (3.42,1.42) circle(2pt)[fill=black] (.58,-1.42) circle(2pt)[fill=black];
 \draw (2,0) -- node[above]{4} (0,0) (2,0) --node[above]{2} (1.1,1.78) --node[above]{2} (0,2) (1.1,1.78) --node[above]{1} (0,4) (2,0) --node[above]{1} (3.42,1.42) (2,0) --node[above]{1} (.58,-1.42) (3.42,1.42) --node[above]{5} (0,6) (.58,-1.42) --node[above]{3} (0,-2);
 \draw[red] (2,0) circle(2);
 }
\end{figure}
